---
title: "Cardiovascular phenotypes to diseases and genes"
output: html_document
---

Want to find genetic diseases (for which the causative gene is known) that have a cardiovascular phenotype.

Starting with "Abnormality of the cardiovascular system" (HP:0001626), download disease associations from Monarch (presumably imported from HPO), save to data table

```{r}
library(data.table)
dt.monarch <- fread("https://solr.monarchinitiative.org/solr/golr/select?defType=edismax&qt=standard&indent=on&wt=csv&rows=100000&start=0&fl=subject,subject_label,relation,relation_label,object,object_label,evidence,evidence_label,source,is_defined_by,qualifier&facet=true&facet.mincount=1&facet.sort=count&json.nl=arrarr&facet.limit=25&facet.method=enum&csv.encapsulator=&csv.separator=%09&csv.header=true&csv.mv.separator=%7C&fq=object_closure:%22HP:0001626%22&fq=subject_category:%22disease%22&facet.field=subject_taxon_closure_label&facet.field=object_taxon_closure_label&q=*:*")

# add new column just with OMIM ID
dt.monarch[,subjectOMIM:=sub("OMIM:","",dt.OMIM$subject)]
dim(dt.monarch)

#filter by diseases mapped to OMIM
setkeyv(dt.monarch,c("subject","object_label"))
dt.monarch <- dt.monarch[like(subject,"OMIM:")]
dim(dt.monarch)

# remove ECO:0000501 (IEA) 
dt.monarch <- dt.monarch[evidence != "ECO:0000501"]
dim(dt.monarch)

# and sort by most common phenotype
print(dt.monarch[,.N,by="object_label"][order(-N)],topn=100)
```

download and parse OMIM file

```{r}
library(yaml)
config <- yaml.load_file("config.yml")
omimURL <- paste("http://data.omim.org/downloads/",config$omimkey,"/morbidmap.txt",sep="")
dt.omim <- fread(omimURL)
dim(dt.omim)

# remove lines that don't have a OMIM ID in the "# Phenotype" column
dt.omim <- dt.omim[grep(' \\d{6} ',dt.omim$`# Phenotype`),]
dim(dt.omim)

# parse out the OMIM Phenotype ID from the '# Phenotype' column into new 'DiseaseID' column
dt.omim[, DiseaseID:=sub('.*(\\d{6}).*','OMIM:\\1',dt.omim$`# Phenotype`)]

# remove duplicated tuples 'MIM Number' - 'DiseaseID'
setkeyv(dt.omim,c("MIM Number","DiseaseID"))
dt.omim <- unique(dt.omim)
dim(dt.omim)

# concatenate genes into a single field --> dt.omim2
dt.omim2 <- dt.omim[,.(concat=paste0(`MIM Number`,collapse=",")),by=.(DiseaseID)]
colnames(dt.omim2)[2] <- "OMIMGeneIDs"
```

join the HPO table and OMIM table into dt.merge
```{r}
setkeyv(dt.monarch,"subject")
setkeyv(dt.omim2,"DiseaseID")
dt.merge <- merge(dt.monarch,dt.omim2,by.x="subject",by.y="DiseaseID")

```

rank phenotypes by # of linked diseases
```{r}
head(dt.merge[,.N,by="object_label"][order(-N)],n=20)

# to investigate the linked diseases for any given phenotype:
dt.merge[object_label=="Ventricular septal defect",.(subject,subject_label,OMIMGeneIDs)]
```

rank genes by # of linked diseases
```{r}
dt.merge2 <- unique(dt.merge[,.(subject,subject_label,OMIMGeneIDs)])

head(dt.merge2[,.N,by="OMIMGeneIDs"][order(-N)],n=20)

# to investigate the linked phenotypes for a given gene ID (eg 150330)
dt.merge2[OMIMGeneIDs=="150330"]

```

